---
title: "Percipitation Concentration in Austin"
author: "Alexander Mohn"
date: "November 13, 2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Having grown up in Austin, TX, I've anecdotally noted that the percipitation patterns seem to be changing here since I was a little kid.
Especially since the drought of 2009-2011, it's seemed to me that our summers have been getting drier and that we're alternating more between drought conditions and flooding.
When I was young, it seemed like we had afternoon thunderstorms on a good number of days during the summer and it seemed like we always had a modest amount of moisture.
Now, it seems like summers have become bone-dry and the rain we do get comes in the form of torrential storms that drop a huge amount of rain in very little time.
Because memory is not super reliable, I decided to investigate this using data from NOAA.

I'm primarily going to focus on the following two questions:

1. Has the concentration of percipitation increased in Austin? In other words, has the amount of percipitation that falls on a rainy day increased?
2. Has the seasonality of percipitation changed in Austin? Are some seasons wetter/drier than they used to be?

But I've got a few more things I'm also interested in investigating:

- Has the concentration of percipitation in each season changed?
- Based on evapotranspiration, does Austin have a more pronounced dry/wet seasonality now?

Time to get started. The preliminaries:
```{r}
# Libraries
library(httr)
library(dplyr)
library(jsonlite)
library(tidyr)
library(R.cache)

authToken <- paste(readLines("authToken.txt"), collapse = "")
```
Find climate stations within .25 degrees of Austin with good data coverage and at least 10 years of data:
```{r}
baseUrl <- "https://www.ncdc.noaa.gov/cdo-web/api/v2/stations"
request <- GET(baseUrl, query=list(extent="30.0172,-97.9931,30.5172,-97.4931",datasetid="GHCND",limit="1000"),add_headers(token=authToken))
json <- content(request, as="text")
if(validate(json))
  json <- fromJSON(json)
  stations <- json$results %>% filter(datacoverage > .8) %>% mutate(maxdate = as.Date(maxdate), mindate = as.Date(mindate)) %>%
    mutate(length = maxdate - mindate) %>% filter(length > 3650) %>% select(-length) %>% 
    mutate(startYear = format(mindate, "%Y"), endYear = format(maxdate, "%Y")) %>% select(-elevation, -elevationUnit)
```
Fetch percipitation data for the stations for the years 1895-2015.
The this code lies in a separate document, to keep this one clutter-free, but you can see it over at the Github repo (add link).
```{r}
source("./getData.R")
getAllData(stations)
```
