---
title: "Percipitation Concentration in Austin"
author: "Alexander Mohn"
date: "November 13, 2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Having grown up in Austin, TX, I've anecdotally noted that the precipitation patterns seem to be changing here since I was a little kid.
Especially since the drought of 2009-2011, it's seemed to me that our summers have been getting drier and that we're alternating more between drought conditions and flooding.
When I was young, it seemed like we had afternoon thunderstorms on a good number of days during the summer and it seemed like we always had a modest amount of moisture.
Now, it seems like summers have become bone-dry and the rain we do get comes in the form of torrential storms that drop a huge amount of rain in very little time.
Because memory is not super reliable, I decided to investigate this using data from NOAA.

I'm primarily going to focus on the following two questions:

1. Has the concentration of precipitation increased in Austin? In other words, has the amount of precipitation that falls on a rainy day increased?
2. Has the seasonality of precipitation changed in Austin? Are some seasons wetter/drier than they used to be?

But I've got a few more things I'm also interested in investigating:

- Has the concentration of precipitation in each season changed?
- Based on evapotranspiration, does Austin have a more pronounced dry/wet seasonality now?

Time to get started. The preliminaries:
```{r}
# Libraries
library(httr)
library(dplyr)
library(jsonlite)
library(tidyr)
library(R.cache)
library(ineq)

authToken <- paste(readLines("authToken.txt"), collapse = "")
```
Find climate stations within .25 degrees of Austin with good data coverage and at least 10 years of data:
```{r}
baseUrl <- "https://www.ncdc.noaa.gov/cdo-web/api/v2/stations"
request <- GET(baseUrl, query=list(extent="30.0172,-97.9931,30.5172,-97.4931",datasetid="GHCND",limit="1000"),add_headers(token=authToken))
json <- content(request, as="text")
if(validate(json))
  json <- fromJSON(json)
  stations <- json$results %>% filter(datacoverage > .8) %>% mutate(maxdate = as.Date(maxdate), mindate = as.Date(mindate)) %>%
    mutate(length = maxdate - mindate) %>% filter(length > 3650) %>% select(-length) %>% 
    mutate(startYear = format(mindate, "%Y"), endYear = format(maxdate, "%Y")) %>% select(-elevation, -elevationUnit)
```
Fetch percipitation data for the stations for the years 1895-2015.
The this code lies in a separate document, to keep this one clutter-free, but you can see it over at the [github repo.](https://github.com/pogrmman/AustinPercipitation)
```{r}
source("./getData.R")
getAllData(stations)
```
I'll start by looking at heavy rainfall days with rainfall in the 90th percentile of all rainfall days (more than 251 mm of rain).
```{r}
# Get all days with rainfall >= trace
rainyDays <- allData %>% filter(value>0|measurement=="T")
deciles <- quantile(rainyDays$value,c(.1,.2,.3,.4,.5,.6,.7,.8,.9))
heavyRainDays <- rainyDays %>% filter(value>deciles["90%"]) %>% separate(date, into=c("year","month","day"), sep="-") %>% 
  mutate(year=as.numeric(year), month=as.numeric(month), day=as.numeric(day)) %>% 
  group_by(year,station,.drop=TRUE) %>% summarize(HeavyRainDays=n()) %>% ungroup() %>% group_by(year) %>% summarize(HeavyRainDays=mean(HeavyRainDays))
heavyRainDays %>% ggplot(aes(year,HeavyRainDays)) + geom_point() + geom_smooth(method="lm") + 
  labs(title = "Number of Heavy Rain Days per Year",
       x = "Year",
       y = "Number of Heavy Rain Days")
fit <- lm(heavyRainDays$HeavyRainDays ~ heavyRainDays$year)
print(deciles)
print(summary(fit))
```
There's not much, if any change in terms of the number of days per year in the 90th percentile for rainfall.
However, if you look at extremely heavy rainfall days (99th percentile -- 775 mm/day), there is a substantial (though non-significant) increase, although the data are extremely noisy and the majority of its variability is not explained by the line of best fit.
```{r}
deciles <- quantile(rainyDays$value, c(.9,.95,.99))
extremeRainDays <- rainyDays %>% filter(value>deciles["99%"]) %>% separate(date, into=c("year","month","day"), sep="-") %>%
  mutate(year=as.numeric(year), month=as.numeric(month), day=as.numeric(day)) %>%
  group_by(year,station,.drop=TRUE) %>% summarize(ExtremeRainDays=n()) %>% ungroup() %>% group_by(year) %>% 
  summarize(ExtremeRainDays = mean(ExtremeRainDays))
extremeRainDays %>% ggplot(aes(year,ExtremeRainDays)) + geom_point() + geom_smooth(method="lm") +
  labs(title = "Extreme Rainfall Days per Year",
       x = "Year",
       y = "Number of Extreme Rainfall Days")
fit <- lm(extremeRainDays$ExtremeRainDays ~ extremeRainDays$year)
print(deciles)
print(summary(fit))
```

```{r}
print(paste("Mean:", as.numeric(mean(rainyDays$value)), "mm"))
print(paste("Median:", as.numeric(median(rainyDays$value)), "mm"))
print(paste("Standard Deviation:", as.numeric(sd(rainyDays$value)), "mm"))
```